---
title: "Palmer Penguins: Does Bill Length Predict Body Mass?"
author: "Jonah Sider-Echenberg"
date: "2026-09-24"
---

## Introduction

Penguins come in different shapes and sizes depending on species, and it's
natural to wonder whether a simple external measurement — bill length — says
anything about a penguin's overall body mass. In this post I explore that
question using the Palmer Penguins dataset.

**Data source:** [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Dr. Kristen Gorman at the Palmer Station Long Term Ecological
Research Network in Antarctica, and distributed as the `palmerpenguins` R
package.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(palmerpenguins)

penguins <- penguins |>
  drop_na(bill_length_mm, body_mass_g, species)
```

The setup chunk above loads the data and drops rows with missing values in
the columns we need. Warnings and messages are suppressed here on purpose —
this chunk is just plumbing, not something a reader needs to see.

## Summary statistics by species

Before looking at any relationship, it's worth checking how bill length and
body mass differ across the three species in the dataset.

```{r}
summary <- penguins |>
  group_by(species) |>
  summarise(
    mean_bill_length_mm = round(mean(bill_length_mm), 1),
    sd_bill_length_mm = round(sd(bill_length_mm), 1),
    mean_body_mass_g = round(mean(body_mass_g), 1),
    sd_body_mass_g = round(sd(body_mass_g), 1)
  )
summary
```

Gentoo penguins stand out immediately as the heaviest species by a wide
margin, while Adelie and Chinstrap are closer to each other in body mass
despite having noticeably different bill lengths.

## Visualizing the relationship

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of penguins by species, plotted against bill length."
penguins_length_to_mass <- ggplot(
    data = penguins,
    mapping = aes(x = bill_length_mm, y = body_mass_g, colour = species)
  ) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Bill length (mm)") +
  ylab("Body mass (g)") +
  labs(colour = "Species") +
  theme_bw() +
  theme(text = element_text(size = 18))

penguins_length_to_mass
```

@fig-penguin-mass shows the scatter plot with a linear trend line fit per
species. The positive slopes confirm that within every species, longer bills go along 
with higher body mass. Gentoo penguins occupy a visibly heavier and longer-billed region 
of the plot than Adelie or Chinstrap.

## Conclusion

Bill length turns out to be a reasonable, if imperfect, predictor of body
mass — both across species and within species. Species membership matters too:
knowing only bill length without knowing species would give an incomplete
picture than looking at the within-species trends shown here.